---
title: "Diversification trial analysis"
author: "Jennifer Schmidt"
date: "3/5/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE,warning=F,message=F}
library(geodist)
library(phyloseq)
library(tidyverse)
library(vegan)
library(ggplot2)
ps16S <- readRDS("tarengge16S.rds")
psITS <- readRDS("tarenggeITS.rds")
shade <- read.csv("data/shade_trees.csv")

head(sample_data(psITS))


controls <- str_subset(sample_data(ps16S)$Local.Label.y,"Control")

ps16S_samples <- subset_samples(ps16S,Group=="Diversification")
psITS_samples <- subset_samples(psITS,Group=="Diversification")

ps16S_controls <- subset_samples(ps16S,Group=="Monoculture")
psITS_controls <- subset_samples(psITS,Group=="Monoculture")
```

## Calculate distance matrix for distance between sampling locations
Distance is in meters - see geodist vignettes.
```{r }
site_coordsB <- sample_data(ps16S) %>% data.frame() %>% select(Longitude,Latitude) 
distgeoB <- geodist(site_coordsB,measure="geodesic") # geodesic distance
max(distgeoB,na.rm=T) # max distance ~115m 
hist(distgeoB)

distgeoB_samples <- geodist(site_coordsB[21:88,],measure="geodesic")
hist(distgeoB_samples)
max(distgeoB_samples,na.rm=T) # only 86m within samples
distgeoB_controls <- geodist(site_coordsB[1:20,],measure="geodesic")
hist(distgeoB_controls) # 80 m

# distance to shade tree - only makes sense to include distances <5m if we assume litterfall, fine roots, etc. only impact the area directly around the tree
sample_data(ps16S)$Dist_shade <- geodist(data.frame(Longitude=sample_data(ps16S)$Longitude,Latitude=sample_data(ps16S)$Latitude),data.frame(Longitude=sample_data(ps16S)$Shade_Longitude,Latitude=sample_data(ps16S)$Shade_Latitude),paired=TRUE,measure="geodesic")

hist(sample_data(ps16S)$Dist_shade)


# Fungi
site_coordsF <- sample_data(psITS) %>% data.frame() %>% rownames_to_column("rownames") %>% select(rownames,Longitude,Latitude) %>%arrange(rownames)%>%column_to_rownames("rownames") # need to arrange by sample name to have controls at the top
distgeoF <- geodist(site_coordsF,measure="geodesic") # geodesic distance
max(distgeoF,na.rm=T) # checking that it's the same
distgeoF_samples <- geodist(site_coordsF[21:99,],measure="geodesic")
hist(distgeoF_samples)
max(distgeoF_samples,na.rm=T) # 86m within samples

distgeoF_controls <- geodist(site_coordsF[1:20,],measure="geodesic")
hist(distgeoF_controls)

sample_data(psITS)$Dist_shade <- geodist(data.frame(Longitude=sample_data(psITS)$Longitude,Latitude=sample_data(psITS)$Latitude),data.frame(Longitude=sample_data(psITS)$Shade_Longitude,Latitude=sample_data(psITS)$Shade_Latitude),paired=TRUE,measure="geodesic")
hist(sample_data(psITS)$Dist_shade)
```

## Bray-Curtis dissimilarity matrices
```{r}
distbrayB=vegdist(data.frame(otu_table(ps16S)), distance="bray")%>%as.matrix()
distjaccardB = vegdist(data.frame(otu_table(ps16S)),distance="jaccard")%>%as.matrix()

distbrayF=vegdist(data.frame(otu_table(psITS)), distance="bray")%>%as.matrix()
distjaccardF = vegdist(data.frame(otu_table(psITS)),distance="jaccard")%>%as.matrix()

distbrayB_samples=vegdist(data.frame(otu_table(ps16S_samples)), distance="bray")%>%as.matrix()
distbrayF_samples=vegdist(data.frame(otu_table(psITS_samples)), distance="bray")%>%as.matrix()

distbrayB_controls=vegdist(data.frame(otu_table(ps16S_controls)), distance="bray")%>%as.matrix()
distbrayF_controls=vegdist(data.frame(otu_table(psITS_controls)), distance="bray")%>%as.matrix()
```


## Bray-Curtis dissimilarity vs. distance
To assess distance-decay relationship (i.e. does dissimilarity between microbial communities increase with increasing geographic distance), we want to plot Bray-Curtis distance vs. geographic distance and fit a trendline.

```{r,echo=FALSE}
#Bacteria - diversification
distB_samples <- data.frame(Bray=as.vector(distbrayB_samples[upper.tri(distbrayB_samples)]), Geo=as.vector(distgeoB_samples[upper.tri(distgeoB_samples)]),Group="Diversification") 
distB_samples %>% ggplot(aes(x=Geo,y=Bray))+geom_point()+geom_smooth(method="lm")+theme_minimal()+theme(panel.grid=element_blank())+labs(x="Geographic distance (m)",y="Bray-Curtis distance",title="Bacteria")+theme(plot.title=element_text(hjust=0.5))

#Fungi - diversification
distF_samples <- data.frame(Bray=as.vector(distbrayF_samples[upper.tri(distbrayF_samples)]), Geo=as.vector(distgeoF_samples[upper.tri(distgeoF_samples)]),Group="Diversification") 
distF_samples %>% ggplot(aes(x=Geo,y=Bray))+geom_point()+geom_smooth(method="lm")+theme_minimal()+theme(panel.grid=element_blank())+labs(x="Geographic distance (m)",y="Bray-Curtis distance",title="Fungi")+theme(plot.title=element_text(hjust=0.5))


# Separate trendlines for diversification and monoculture
distB_controls <- data.frame(Bray=as.vector(distbrayB_controls[upper.tri(distbrayB_controls)]), Geo=as.vector(distgeoB_controls[upper.tri(distgeoB_controls)]),Group="Monoculture")
distB_all = rbind(distB_samples,distB_controls)
distB_all %>% ggplot(aes(x=Geo,y=Bray))+geom_point(aes(color=Group),alpha=0.2)+geom_smooth(aes(group=Group,color=Group),method="lm",alpha=0.2)+theme_minimal()+theme(panel.grid=element_blank())+labs(x="Geographic distance (m)",y="Bray-Curtis distance",title="Bacteria")+theme(plot.title=element_text(hjust=0.5))+scale_color_brewer(palette="Set2")+theme(legend.position="bottom")


distF_controls <- data.frame(Bray=as.vector(distbrayF_controls[upper.tri(distbrayF_controls)]), Geo=as.vector(distgeoF_controls[upper.tri(distgeoF_controls)]),Group="Monoculture") 
distF_all <- rbind(distF_samples,distF_controls)
distF_all %>% ggplot(aes(x=Geo,y=Bray))+geom_point(aes(color=Group),alpha=0.2)+geom_smooth(aes(group=Group,color=Group),method="lm",alpha=0.2)+theme_minimal()+theme(panel.grid=element_blank())+labs(x="Geographic distance (m)",y="Bray-Curtis distance",title="Fungi")+theme(plot.title=element_text(hjust=0.5))+scale_color_brewer(palette="Set2")+theme(legend.position="bottom")
```

# Correlations
```{r}
cor(distB_all$Bray,distB_all$Geo,use="complete.obs") # 0.422
cor(distF_all$Bray,distF_all$Geo,use="complete.obs") # 0.146
```

# Regression equation
```{r}
modB_samples <- lm(Bray ~ Geo, data = distB_samples)
summary(modB_samples)# slope 2.248e-03,p<2e-16 ***
modB_controls <- lm(Bray ~ Geo, data = distB_controls)
summary(modB_controls)# slope 3.617e-04,p=1.34e-15 ***

modF_samples <- lm(Bray ~ Geo, data = distF_samples)
summary(modF_samples) # slope 6.683e-04 int 6.846e-01, p<0.001
modF_controls <- lm(Bray~Geo,data = distF_controls)
summary(modF_controls)
```

# Multiple matrix regression analysis (MMRR)
Wang et al. (2013) proposed this as an alternative to Mantel tests, which are increasingly thought of as less suitable for genetic/geographic analyses.
```{r}
# define functions as per Wang et al.
MMRR<-function(Y,X,nperm=999){
	#compute regression coefficients and test statistics
	nrowsY<-nrow(Y)
	y<-unfold(Y)
	if(is.null(names(X)))names(X)<-paste("X",1:length(X),sep="")
        Xmats<-sapply(X,unfold)
        fit<-lm(y~Xmats)
	coeffs<-fit$coefficients
	summ<-summary(fit)
	r.squared<-summ$r.squared
	tstat<-summ$coefficients[,"t value"]
	Fstat<-summ$fstatistic[1]
	tprob<-rep(1,length(tstat))
	Fprob<-1

	#perform permutations
	for(i in 1:nperm){
		rand<-sample(1:nrowsY)
		Yperm<-Y[rand,rand]
		yperm<-unfold(Yperm)
		fit<-lm(yperm~Xmats)
		summ<-summary(fit)
                Fprob<-Fprob+as.numeric(summ$fstatistic[1]>=Fstat)
                tprob<-tprob+as.numeric(abs(summ$coefficients[,"t value"])>=abs(tstat))
	}

	#return values
	tp<-tprob/(nperm+1)
	Fp<-Fprob/(nperm+1)
	names(r.squared)<-"r.squared"
	names(coeffs)<-c("Intercept",names(X))
	names(tstat)<-paste(c("Intercept",names(X)),"(t)",sep="")
	names(tp)<-paste(c("Intercept",names(X)),"(p)",sep="")
	names(Fstat)<-"F-statistic"
	names(Fp)<-"F p-value"
	return(list(r.squared=r.squared,
		coefficients=coeffs,
		tstatistic=tstat,
		tpvalue=tp,
		Fstatistic=Fstat,
		Fpvalue=Fp))
}

# unfold converts the lower diagonal elements of a matrix into a vector
# unfold is called by MMRR

unfold<-function(X){
	x<-vector()
	for(i in 2:nrow(X)) x<-c(x,X[i,1:i-1])
	x<-scale(x, center=TRUE, scale=TRUE)  # Comment this line out if you wish to perform the analysis without standardizing the distance matrices! 
	return(x)
}

# Bacteria
#Make a list of the explanatory (X) matrices
XmatsB <- list(geography=distgeoB_samples)#add soil data when received
distgeoB_samples <- as.matrix(distgeoB_samples)
# Run MMRR function using genetic distance as the response variable and Xmats as the explanatory variables.
# nperm does not need to be specified, default is nperm=999)
MMRR(distbrayB_samples,XmatsB,nperm=999) # p<0.01

XmatsB <- list(geography=distgeoB_controls)
MMRR(distbrayB_controls,XmatsB,nperm=999) # p>0.05

# Fungi
XmatsF <- list(geography=distgeoF_samples)
MMRR(distbrayF_samples,XmatsF,nperm=999) # NS
XmatsF <- list(geography=distgeoF_controls)
MMRR(distbrayF_controls,XmatsF,nperm=999) # NS
```


# PCoA ordination of microbial communities
```{r}
# PCoA based on Bray-Curtis dissimilarity
ps16Slog <- transform_sample_counts(ps16S, function(x) log(1 + x))
out.pcoa.logB <- ordinate(ps16Slog,  method = "PCoA", distance = "bray")
evalsB <- out.pcoa.logB$values[,1]

(comp16S <- plot_ordination(ps16Slog, out.pcoa.logB, color = "Group") +
  labs(col = "Trial")+ theme_bw()+
  coord_fixed(sqrt(evalsB[2] / evalsB[1]))+
  theme(text=element_text(size=14),axis.title=element_text(size=14),legend.title=element_text(size=14),plot.title=element_text(hjust=0.5))+
  scale_color_brewer(palette="Set2")+
  theme(panel.grid=element_blank())+theme(legend.position="bottom")+theme(plot.margin=margin(1,1,1,1))+ggtitle("Bacteria"))

# Fungi
psITSlog <- transform_sample_counts(psITS, function(x) log(1 + x))
out.pcoa.logF <- ordinate(psITSlog,  method = "PCoA", distance = "bray")
evalsF <- out.pcoa.logF$values[,1]

(compITS <- plot_ordination(psITSlog, out.pcoa.logF, color = "Group") +
  labs(col = "Trial")+ theme_bw()+
  coord_fixed(sqrt(evalsF[2] / evalsF[1]))+
  theme(text=element_text(size=14),axis.title=element_text(size=14),legend.title=element_text(size=14),plot.title=element_text(hjust=0.5))+
  scale_color_brewer(palette="Set2")+
  theme(panel.grid=element_blank())+theme(legend.position="bottom")+theme(plot.margin=margin(1,1,1,1))+ggtitle("Fungi"))
```

# Effect of shade trees
```{r}
# only use samples where shade tree coordinates are known and shade tree is within 7 m (generous - reduces to 24 samples)
head(sample_data(ps16S_samples))
ps16S_shade <- subset_samples(ps16S_samples,complete.cases(Shade_Latitude)&Dist_shade<7)

ps16Slog_samples <- transform_sample_counts(ps16S_shade, function(x) log(1 + x))
out.pcoa.logB_samples <- ordinate(ps16Slog_samples,  method = "PCoA", distance = "bray")
evalsB_samples <- out.pcoa.logB_samples$values[,1]
(shade16S <- plot_ordination(ps16Slog_samples, out.pcoa.logB_samples, color = "Shade_species.y") +
  labs(col = "Nearest Shade Species")+ theme_bw()+
  coord_fixed(sqrt(evalsB_samples[2] / evalsB_samples[1]))+
  theme(text=element_text(size=14),axis.title=element_text(size=14),legend.title=element_text(size=14))+
  scale_color_brewer(palette="BrBG")+
  theme(panel.grid=element_blank())+theme(plot.margin=margin(1,1,1,1))+ggtitle("Bacteria")+theme(legend.position = "bottom")+theme(plot.title = element_text(hjust=0.5))+guides(col=guide_legend(ncol=2)))

#Would analyze with ANOSIM, which is robust to unequal group sizes unlike PERMANOVA, but we don't have enough members of each group
#anosim(data.frame(otu_table(ps16S_shade)), sample_data(ps16S_shade)$Shade_species.y, permutations = 999, distance = "bray", strata = NULL) # p<0.05 for both bray and jaccard
 
# Fungi
psITSlog_samples <- transform_sample_counts(psITS_samples, function(x) log(1 + x))
out.pcoa.logF_samples <- ordinate(psITSlog_samples,  method = "PCoA", distance = "jaccard")
evalsF_samples <- out.pcoa.logF_samples$values[,1]
(shadeITS <- plot_ordination(psITSlog_samples, out.pcoa.logF_samples, color = "Shade_species") +
  labs(col = "Nearest Shade Species")+ theme_bw()+
  coord_fixed(sqrt(evalsF_samples[2] / evalsF_samples[1]))+
  theme(text=element_text(size=14),axis.title=element_text(size=14),legend.title=element_text(size=14))+
  scale_color_brewer(palette="BrBG")+
  theme(panel.grid=element_blank())+theme(plot.margin=margin(1,1,1,1))+ggtitle("Fungi")+theme(legend.position = "bottom")+theme(plot.title = element_text(hjust=0.5))+guides(col=guide_legend(ncol=2)))

#ANOSIM - robust to unequal group sizes, unlike permanova
anosim(data.frame(otu_table(psITS_samples)), sample_data(psITS_samples)$Shade_species, permutations = 999, distance = "bray", strata = NULL) # NS for both
```


# Overall alpha diversity
```{r}
library(microbiomeutilities)
#richness
sample_data(psITS)$Richness <- specnumber(otu_table(psITS),MARGIN=1) # 1=rows, 2=col; samples are rows
# shannon index
sample_data(psITS)$Shannon <- vegan::diversity(otu_table(psITS), "shannon")
sample_data(psITS)$Pielou <- (sample_data(psITS)$Shannon/log(sample_data(psITS)$Richness))


div <- data.frame(sample_data(psITS)) %>% select("ID","Group","Richness","Shannon","Pielou") %>%
  rename("Shannon Index" = "Shannon","Pielou Index" = "Pielou") %>% pivot_longer(cols=c("Richness","Shannon Index","Pielou Index"),names_to="div")

(divITS <- ggplot(div,aes(x=Group,y=value))+
  geom_point(aes(col=Group))+theme_bw()+
  facet_wrap(~div,scales="free")+
  scale_color_brewer(palette="Set2")+
  theme(panel.grid=element_blank())+
  theme(text=element_text(size=14))+
  theme(axis.text.x=element_blank(),axis.title.x=element_blank(),axis.ticks.x=element_blank())+
  theme(axis.title.y=element_blank())+theme(legend.position="bottom"))


# Bacteria
#richness
sample_data(ps16S)$Richness <- specnumber(otu_table(ps16S),MARGIN=1)
# shannon index
sample_data(ps16S)$Shannon <- vegan::diversity(otu_table(ps16S), "shannon")
sample_data(ps16S)$Pielou <- (sample_data(ps16S)$Shannon/log(sample_data(ps16S)$Richness))


div <- data.frame(sample_data(ps16S)) %>% select("ID","Group","Richness","Shannon","Pielou") %>%
  rename("Shannon Index" = "Shannon","Pielou Index" = "Pielou") %>% pivot_longer(cols=c("Richness","Shannon Index","Pielou Index"),names_to="div")

(div16S <- ggplot(div,aes(x=Group,y=value))+
  geom_point(aes(col=Group))+theme_bw()+
  facet_wrap(~div,scales="free")+
  scale_color_brewer(palette="Set2")+
  theme(panel.grid=element_blank())+
  theme(text=element_text(size=14))+
  theme(axis.text.x=element_blank(),axis.title.x=element_blank(),axis.ticks.x=element_blank())+
  theme(axis.title.y=element_blank())+theme(legend.position="bottom"))
```

# Overall compositions: Diversification vs. Monoculture Plots
```{r}
library(microViz)
(comp16S <- ps16S %>%phyloseq::merge_samples(group = "Group")%>% tax_fix %>%
  comp_barplot(
    tax_level = "Phylum", n_taxa = 12,
    bar_outline_colour = "grey5",
    merge_other = FALSE,
    sample_order = "bray", 
    tax_order = sum,
      bar_width = 0.7) + coord_flip() +
    theme(legend.position="bottom")+theme(axis.text=element_text(size=10),axis.title=element_text(size=10),strip.text=element_text(size=10),legend.text=element_text(size=10))+ guides(fill=guide_legend(ncol=3))+ theme(legend.title = element_blank())+theme(plot.margin = margin(0, 0.6, 0, 0.5, "cm")))

(compITS <- psITS %>%phyloseq::merge_samples(group = "Group")%>% tax_fix %>%
  comp_barplot(
    tax_level = "Phylum", n_taxa = 12,
    bar_outline_colour = "grey5",
    merge_other = FALSE,
    sample_order = "bray", 
    tax_order = sum,
      bar_width = 0.7) + coord_flip() +
    theme(legend.position="bottom")+theme(axis.text=element_text(size=10),axis.title=element_text(size=10),strip.text=element_text(size=10),legend.text=element_text(size=10))+ guides(fill=guide_legend(ncol=3))+ theme(legend.title = element_blank())+theme(plot.margin = margin(0, 0.5, 0, 0.5, "cm")))
```


